---
title: "WGCNA"
author: "Torgeir R. Hvidsten & Dorota Kawa"
date: "31/10/2019"
output:
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(WGCNA)
options(stringsAsFactors = FALSE)
library(tidyverse)
library(gplots)

library(DT)
library(ggbiplot)

```

## Load and filter data

```{r filter, warning=FALSE, message=FALSE}

ext <- "unsigned_0.35"
compute = FALSE

# Load the expression data from ATLAS, FIELD and POT experiment (including MCO)
counts <- read.table(file="/Users/dorota/Dropbox/UC_DAVIS/FIELD_SAMPLES/2019/Counts-and-tpm/20191120-Atlas-Field-Pot-TPM-quantile-norm-together-withMCO.csv", header=TRUE, row.names=1, sep=",")
head(counts)
dim(counts)
# Select most variable genes 
counts$variance <- apply(counts, 1, var)
counts2 <- counts[counts$variance >= quantile(counts$variance, c(.25)), ] #75% most variable genes
counts2$variance <- NULL

# log2 transformation
data.tmp <- counts2[which(apply(counts2, 1, function(x) mean(x)>1)),]
data <- log2(data.tmp+1)

paste0("Genes x Samples: ", paste0(dim(data), collapse = " x "))
plot(density(as.matrix(data)), xlab = "Gene expression values", ylab = "", main = "", col = "red", lwd=5)
dat <- data

dat %>%
  gather(Samples, Expression) %>%
  ggplot(aes(x = Expression, col = Samples)) +
  geom_density()
  

```

## PCA/Hierarchical clustering

```{r pca, warning=FALSE}

# PCA
meta <- read.table(file="/Users/dorota/Dropbox/UC_DAVIS/FIELD_SAMPLES/2019/Counts-and-tpm/metadata_field&atlas&potwithMCO.csv", header=TRUE, sep=",")
pc <- prcomp(t(dat))
var.expl <- pc$sdev^2 / sum(pc$sdev^2)
paste0("Varance explained: ", paste0(format(var.expl[1:5], digits = 2), collapse = " "))
samples <- colnames(dat)

ggbiplot(pc,labels=meta$Promoter, var.axes = FALSE, groups=meta$Promoter)
ggplot(as.data.frame(pc$x), aes(PC1, PC2, shape = meta$Experiment, col = meta$Promoter)) + 
geom_point()+
  xlab(paste0("PC1 (", round(var.expl[1], digits=2),")")) +
  ylab(paste0("PC2 (", round(var.expl[2], digits=2),")")) +
  ggtitle("PCA")

# Hierarchical clustering
dist.var <- as.dist(1-cor(dat))
dist.var.tre <- hclust(dist.var, method = "ward.D")
plot(dist.var.tre, cex = 1, sub = "", xlab = "", main = "Hierarchical clustering")

if (FALSE) {
  if (compute) {
    dist.obs <- as.dist(1-cor(t(dat)))
    dist.obs.tre <- hclust(dist.obs, method = "ward.D") 
    save(dist.var.tre, dist.obs.tre, file = paste0("trees", ext, ".RData"))
  } else {
    load(paste0("trees", ext, ".RData"))
  }

  heatmap.2(x = as.matrix(dat), 
            Rowv = as.dendrogram(dist.obs.tre),
            Colv = as.dendrogram(dist.var.tre),
            trace = "none",
            scale = "row",
            col = colorRampPalette(c("blue","white","red")),
            labRow = c(""),
            labCol = samples,
            cexCol = 0.75,
            srtCol = 45
  )
}

```

## Soft thresholding

Find the soft threshold power for the WGCNA network. 

```{r soft_thr, warning=FALSE}
dat_t <- t(dat)

powers = c(c(1:10), seq(from = 12, to=20, by=2))
if (compute) {
  sft = pickSoftThreshold(dat_t, powerVector = powers, corFnc = "bicor", networkType = "unsigned")
  save(sft, file = paste0("sft", ext, ".RData"))
} else {
  load(paste0("sft", ext, ".RData"))
}
# Plot the results
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.80,col="red")

power <- 5
```

## Network module finding

And some plots to show the similarity within and between modules.

```{r modules, warning=FALSE}
if (compute) {
  modules <- blockwiseModules(datExpr = dat_t,
                            power = power, 
                            networkType = "unsigned", 
                            corType = "bicor",
                            maxPOutliers = 0.05,
                            TOMType = "signed",
                            mergeCutHeight = 0.35,
                            maxBlockSize = 20000,
                            saveTOMs = TRUE,
                            saveTOMFileBase = paste0("TOM", ext)
                            )
  save(modules, file = paste0("modules", ext, ".RData"))
  
  load(paste0("TOM", ext, "-block.1.RData"))
  TOM <- as.matrix(TOM)
  save(TOM, file = paste0("TOM", ext, "-block.1.RData"))
} else {
  load(paste0("modules", ext, ".RData"))
}

paste0("Samples x Modules: ", paste0(dim(modules$MEs), collapse = " x "))

# If TOM is needed
#load(paste0("TOM", ext, "-block.1.RData"))

#mergedColors = labels2colors(modules$colors)
#plotDendroAndColors(modules$dendrograms[[1]], 
#                    mergedColors[modules$blockGenes[[1]]],
#                    "Module colors",
#                    dendroLabels = FALSE, hang = 0.03,
#                    addGuide = TRUE, guideHang = 0.05)

MEs <- modules$MEs

kME <- bicor(dat_t, MEs, maxPOutliers = 0.05)

# Correlation within modules
intra_cor <- c()
for (i in 1:nrow(dat)) {
  m <- modules$colors[i]
  intra_cor[i] <- kME[i, paste0("ME", m)]
}
hist(intra_cor, xlim = c(-1,1), breaks = seq(-1,1,0.1),
     main = "Correlations with module eigengene (within module correlation)",
     xlab = "Correlation")

# Correlation between modules
MEs_R <- bicor(MEs, MEs, maxPOutliers = 0.05)
hist(MEs_R[upper.tri(MEs_R)], xlim = c(-1,1), breaks = seq(-1,1,0.1),
     main = "Correlations of module eigengenes (between module correlation)",
     xlab = "Correlation")

# Reassign genes to modules (if used, can be stricter than the WGCNA reassignment)
R_th <- 0.7
if (FALSE) {
  c <- gsub("ME", "", colnames(MEs))
  new_color <- rep("darkgrey",length(modules$colors))
  max_sim <- apply(kME, 1, max)
  for (i in 1:ncol(MEs)) {
    new_color[kME[,i] == max_sim & kME[,i] > R_th] = c[i]
  }
  names(new_color) <- names(modules$colors)
  modules$colors <- new_color
}

aa <-as.data.frame(modules$colors)

```

## Network degrees

Compute different network connectivity (degree) measures.

```{r centrality, warning=FALSE}

# Compute degree
if (compute) {
  degrees <- intramodularConnectivity.fromExpr(dat_t, colors = modules$colors, power = power,
                                             networkType = "signed", distFnc = "bicor")
  save(degrees, file = paste0("degrees", ext, ".RData"))
} else {
  load(paste0("degrees", ext, ".RData"))
}

plot (density(degrees$kTotal[!is.na(degrees$kTotal)]), main = "Degree: Full network (kTotal)", xlab = "Degree")
abline(v = 0, col = "red")
plot (density(degrees$kWithin[!is.na(degrees$kWithin)]), main = "Degree: Within module (kWithin)", xlab = "Degree")
abline(v = 0, col = "red")
plot (density(degrees$kDiff[!is.na(degrees$kDiff)]), main = "Degree: Within minus outside module (kDiff)", xlab = "Degree")
abline(v = 0, col = "red")
```

## Modules - Expression profiles

Expression profile plots for each module.

```{r spec_modules, warning=FALSE}

met_mod <- modules$colors[-grep("Solyc", rownames(dat))]
IntGenes <- data.frame(ID = rownames(dat)[1:10])

annot <- read_delim("/Users/dorota/Dropbox/UC_DAVIS/FIELD_SAMPLES/2019/WGCN/SolycAnnotationsITAG310.csv", delim = ",", col_types = cols())
if (FALSE) { # Additional interesting functions
  hyp <- c("carboxylase","reductase","peroxidase","oxidase","laccase")
  for (f in hyp) {
    idx <- grep(f, annot$`arabi-defline`)
    new_genes <- unique(annot$locusName[idx])
    idx <- match(new_genes, IntGenes$ID)
    idx <- which(!is.na(idx))
    if (length(idx) > 0) {
      new_genes <- new_genes[-idx]
    }
    if (length(new_genes) > 0) {
      IntGenes <- rbind(IntGenes,data.frame(ID = new_genes, Symbol = f))
    }
  }
}


# Look at each module
means_df <- data.frame(row.names = rownames(dat), 
                       EP_Atlas=rowMeans(dat[,which(meta$DK_name=="EP_Atlas")]),
                       EXO_Atlas=rowMeans(dat[,which(meta$DK_name=="EXO_Atlas")]),
                       COR_Atlas=rowMeans(dat[,which(meta$DK_name=="COR_Atlas")]),
                       EN_Atlas=rowMeans(dat[,which(meta$DK_name=="EN_Atlas")]),
                       X35S_Atlas=rowMeans(dat[,which(meta$DK_name=="X35S_Atlas")]),
                       MZ_Atlas=rowMeans(dat[,which(meta$DK_name=="MZ_Atlas")]),
                       ACT_Atlas=rowMeans(dat[,which(meta$DK_name=="ACT_Atlas")]),
                       PH_Atlas=rowMeans(dat[,which(meta$DK_name=="PH_Atlas")]),
                       XY_Atlas=rowMeans(dat[,which(meta$DK_name=="XY_Atlas")]),
                       MCO_Atlas=rowMeans(dat[,which(meta$DK_name=="MCO_Atlas")]),
                       V_Atlas=rowMeans(dat[,which(meta$DK_name=="V_Atlas")]),
                       WOX_Atlas=rowMeans(dat[,which(meta$DK_name=="WOX_Atlas")]),
                       EN_Field=rowMeans(dat[,which(meta$DK_name=="EN_Field")]),
                       MCO_Field=rowMeans(dat[,which(meta$DK_name=="MCO_Field")]),
                       MZ_Field=rowMeans(dat[,which(meta$DK_name=="MZ_Field")]),
                       X35S_Field=rowMeans(dat[,which(meta$DK_name=="X35S_Field")]),
                       COR_LR_Pot=rowMeans(dat[,which(meta$DK_name=="COR_LR_Pot")]),
                       COR_AR_Pot=rowMeans(dat[,which(meta$DK_name=="COR_AR_Pot")]),
                       MCO_LR_Pot=rowMeans(dat[,which(meta$DK_name=="MCO_LR_Pot")]),
                       MCO_AR_Pot=(dat[,which(meta$DK_name=="MCO_AR_Pot")])
)
dat<-means_df

eigens_means <-   data.frame(row.names = rownames(t(MEs)), 
                       EP_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="EP_Atlas")]),
                       EXO_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="EXO_Atlas")]),
                       COR_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="COR_Atlas")]),
                       EN_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="EN_Atlas")]),
                       X35S_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="X35S_Atlas")]),
                       MZ_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="MZ_Atlas")]),
                       ACT_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="ACT_Atlas")]),
                       PH_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="PH_Atlas")]),
                       XY_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="XY_Atlas")]),
                       MCO_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="MCO_Atlas")]),
                       V_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="V_Atlas")]),
                       WOX_Atlas=rowMeans(t(MEs)[,which(meta$DK_name=="WOX_Atlas")]),
                       EN_Field=rowMeans(t(MEs)[,which(meta$DK_name=="EN_Field")]),
                       MCO_Field=rowMeans(t(MEs)[,which(meta$DK_name=="MCO_Field")]),
                       MZ_Field=rowMeans(t(MEs)[,which(meta$DK_name=="MZ_Field")]),
                       X35S_Field=rowMeans(t(MEs)[,which(meta$DK_name=="X35S_Field")]),
                       COR_LR_Pot=rowMeans(t(MEs)[,which(meta$DK_name=="COR_LR_Pot")]),
                       COR_AR_Pot=rowMeans(t(MEs)[,which(meta$DK_name=="COR_AR_Pot")]),
                       MCO_LR_Pot=rowMeans(t(MEs)[,which(meta$DK_name=="MCO_LR_Pot")]),
                       MCO_AR_Pot=(t(MEs)[,which(meta$DK_name=="MCO_AR_Pot")])
)
  
MEs <- t(eigens_means)

plots <- c()
nm <- 1
no_mod <- length(unique(modules$colors))
for (col in unique(modules$colors)) {
  
  mod <- rownames(dat)[modules$colors == col]
  met <- names(met_mod[met_mod==col])
  
  layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), widths=c(1,1), heights=c(3,2))

  # Plot expression profiles
  center <- MEs[,paste0("ME", col)]
  center <- scale(center)
  
  no.samples <- ncol(dat)
  
  par(xpd = T, mar = c(4,4,4,12))
  plot(1:no.samples, 
       center[1:no.samples],
       xlab = "", ylab = "Expression (Scaled)", 
       type="b", 
       lty=1, pch=20, lwd=4, xaxt="n", yaxt="n",
       main = paste0("Module ",col," (", nm, "/", no_mod, "): ", length(mod), " genes"),
       col="black",
       ylim = range(scale(t(dat[modules$colors == col,])))
  )
  
  l <- colnames(dat)
  l <- gsub("SQR_", "", l)
  
  axis(1, at=1:no.samples, labels = FALSE)
  text(x = 1:no.samples, par("usr")[3] - 0.5, labels = l, 
       srt = 90, pos = 1, xpd = TRUE, cex=0.8)
  axis(2, cex.axis=1)
  box(col=col, lwd=6)
  
  corr <- c()
  
  idx <- match(mod, rownames(dat))
  for (j in idx) {
    g <- t(dat[j,])
    g <- scale(g)
    corr <- c(corr, cor(center, g))
    lines(1:no.samples, 
         g[1:no.samples],
         col = "lightgrey")
  }
  
  lines(1:no.samples, center[1:no.samples], type="b", lty=1, pch=20, lwd=4, col="black")
  
  m_col <- c("blue","red","green","orange","pink")
  for (j in 1:length(met)) {
    m <- t(dat[rownames(dat) == met[j],])
    m <- scale(m)
    lines(1:no.samples, m[1:no.samples], lwd=2, col = m_col[j])
  }
  
  legend(no.samples + 0.5,1,c("Eigengene", met), lwd = c(4,rep(2, length(met))), 
         col = c("black", m_col[1:length(met)]), lty=1, cex = 0.8)

  par(mar=c(4, 4, 1, 1))
  
  # Plot expression similarity
  hist(corr, xlab = "Correlation to eigengene", ylab = "", breaks = seq(-1,1,0.1),
        main = "", lwd = 2, xlim = c(-1,1))
  
  par(mar=c(0, 0, 0, 0))
  
  nm <- nm+1
}

if (FALSE) { # Draw networks
  
  A <- adjacency(dat_t,
                 type = "signed",
                 power = power,
                 corFnc = "bicor"
  )
  
  for (col in c("all", unique(modules$colors))) {
  
    if (col == "all") {
      mod <- rownames(dat)
      met <- names(met_mod[met_mod==col])
    } else {
      mod <- rownames(dat)[modules$colors == col]
      met <- names(met_mod[met_mod==col])
    }
    
    if (length(met) > 0) {
      
      if (col == "all") {
        module_network <- A
      } else {
        idx <- modules$colors == col
        module_network <- A[idx,idx]
        #diag(module_network) <- 0
        rownames(module_network) <- rownames(dat[idx,])
        colnames(module_network) <- rownames(dat[idx,])
      }
    
      I <- IntGenes
      I <- rbind(I,data.frame(ID = met, Symbol = met))
      
      net_idx <- match(I$ID, rownames(module_network))
      int_idx <- which(!is.na(net_idx))
      net_idx <- net_idx[!is.na(net_idx)]
      
      thr <- 0.8
      
      if (FALSE) {
        hist(module_network, xlab = "Adjacency", ylab = "", breaks = seq(0,1,0.05),
           main = "", col = "red", lwd=5)
        abline(v=thr)
        (sum(module_network > thr) - nrow(module_network))/2
      }
      
      altnames <- rep("", nrow(module_network))
      altnames[net_idx] <- I$Symbol[int_idx]
      
      network <- exportNetworkToCytoscape(adjMat = module_network,
                                          nodeFile = paste0("nodes-", col, ".txt"), 
                                          edgeFile = paste0("edges-", col, ".txt"), 
                                          threshold = thr, 
                                          weighted = TRUE,
                                          altNodeNames = altnames,
                                          nodeAttr = modules$colors[idx]
                                          )
    }
  }
}

```

## Modules - Genes information

A table with information for the genes in the network including module membership and network connectivity. Filter on columns to see e.g. genes from one module.

```{r gene_info, warning=FALSE}

idx <- match(rownames(dat), annot$ID)
subannot <- annot[idx,]
subannot$ID <- rownames(dat)

subannot <- cbind(subannot, degrees[,c(1,2,4)])
subannot$kTotal <- format(subannot$kTotal, digits=0)
subannot$kWithin <- format(subannot$kWithin, digits=0)
subannot$kDiff <- format(subannot$kDiff, digits=0)
subannot$Module <- modules$colors

idx <- match(IntGenes$ID, subannot$locusName)
subannot$ManualAnnot <- rep("", nrow(subannot))
#subannot$ManualAnnot[idx] <- IntGenes$Symbol

datatable(subannot, rownames = FALSE, filter = "top")
```

```{r}

write.csv(subannot,file = "20191126-0.35-Modules-withMCO.csv")
```

##Hub Genes

The gene that has the maximum connections with other genes in a module
```{r hub_genes, warning=FALSE}
Hub_genes <- chooseTopHubInEachModule(datExpr = dat, 
                                      colorh = modules,
                                      power = power, 
                                      type = "unsigned",
                                      omitColors = "grey")
  #This doesn't work..yet 
                            
                      
```
